Simple connectome inference from partial correlation statistics in calcium imaging

### 2014-06-27 article.tex
changement dans l'appendice

To filter out small variations in the signal obtained after applying the
function $g$ as well as to eliminate negative values, we use the following
hard-threshold filter
\begin{align}
h(x^{t}_i) &= x^{t}_i \mathbb{1}(x^{t}_i \geq \tau) \text{ with } \tau > 0,
\end{align}
yielding Figure \ref{fig:threshold_curve}.
As we can see, the processed signal only contains clean spikes.

@@

\section{Particularities of ``Full method''}
\label{app:optimized}

This section provides a full description of the method specifically tuned for
the challenge. Since it is no more than an optimized version of the method
presented in Sections~\ref{sec:filter}~to~\ref{sec:results}, the description
will be limited to explaining the differences that exist between this method
and its optimized version.

This appendix is structured as follows: Section~\ref{sapp:signal} describes
the differences about the signal processing methods applied on fluorescent
calcium time series; Section~\ref{sapp:averaging} then provides a detailed
explanation of the averaging approach; Section~\ref{sapp:connectome} presents
a heuristic trying to retrieve directivity from fluorescent calcium time
series while Section~\ref{sapp:results} finally present results validating
some choices that have been made.

@@

In Section~\ref{sec:filter}, we introduce two composed filtering functions in
order (i) to remove noise induced by light scattering artifact, (ii) to take
into account the slow decay of fluorescence signal and (iii) to reduce the
importance of global high activity in the network. However, in order to win the
challenge, we have heavily optimized several different, but yet similar,
filtering composed function. Taken individually, those composed functions in
conjunction with partial correlation might lead to worse AUROC, but improved
whenever they are averaged (see \ref{tab:} ).

Besides filters $f_1$ and $f_2$ respectively defined by equations \ref{eq:symetric-median} 
and \ref{eq:weighted-asymetric-median}, we also consider two
other low-pass filters
\begin{align}
f_3(x^t_i) &= x^{t-1}_i + x^{t}_i + x^{t+1}_i + x^{t+2}_i, \label{eq:asymetric-median-forward} \\
f_4(x^t_i) &=  x_i^t + x^{t+1}_i  + x^{t+2}_i + x^{t+3}_i. \label{eq:asymetric-median}
\end{align}

After the hard thresholded filter $h$, they were significant differences in
peak magnitudes. Since timeseries $x_i^t$  were mainly between 0 and 1,
we smoothened these differences in peak magnitudes through
\begin{align}
p(x^t_i) = (x_i^t)^c
\end{align}
yielding slightly better AUROC with $c=0.9 \in (0, 1)$.

After filter $p$, we applied a filter $w^*$ similar to $w$, whose goal is also
to magnify the importance of spikes that occur in case of low global activity.
Instead of equation \ref{eq:magnify-filter}, we had
\begin{align}
 w^*(x^{t}_i) &= {(x^{t}_i + 1 )^{\left (1 + \frac{1}{\sum_{j} x^{t}_j}\right )}}^{k(\sum_{j} x^{t}_j)}
\end{align}
where the function $k$ is a piecewise linear function optimized for each filter
$f_1$, $f_2$, $f_3$ and $f_4$. By comparison, the filter $w$ used the following
function $k(\sum_{j} x^{t}_j) = 1$. For more details on this aspect, we refer
the reader to the implementation \footnote{available on
\url{https://github.com/asutera/kaggle- connectomics}.}.

@@

The AUROC performance of the conjunction of the composed function $w^* \circ p
\circ g \circ h \circ f_1$ (respectively $f_2$, $f_3$ and $f_4$) and the
partial correlation statistics was quite sensitive to the choice of $\tau$, the
hard threshold value and the design of the low pass filter. Similarly to the
simplified method, we average the partial correlation statistics obtained by
varying $\tau$ linearly from $0.100$ to $0.209$ and the low-pass filter.

Given the unequal AUROC performance of $f_1$, $f_2$, $f_3$ and $f_4$, the
contribution of each filter in the degree of association between two neurons
was chosen as to maximize the AUROC on \textit{normal-1} and is function of
their individual performance. By comparison in the simple method, each filter
$f_1$ and $f_2$ were given equal weight.

\subsection{Connectome inference from pairwise interaction statistics}
\label{sapp:connectome}

Since partial correlation statistics is a symmetric measure and the connectome
is a directed graph, it should be rewarding to detect the edge orientation. In
this section, we will present a heuristic trying to retrieve directivity.
However, considering results in Section \ref{sec:results}, we expect that the
additional gain margin should be extremely limited.

The heuristic is based on the following observation. The rise of fluorescence
of a neuron indicates its activation. If another neuron is activated after a
slight delay, this could have been caused by the activation of the first
neuron and therefore highlight a (direct) directed relationship between the
two neurons. With that in mind, the heuristic principle is to count the number
of such events occurred between neurons $i$ and $j$ for consecutive time
steps. It can be done by counting the number of times that a transition from
neuron $i$ to neuron $j$ for two consecutive time steps, or mathematically we
compute for each pair $(i,j)$
\begin{align}
s_{ij} = \sum_{t=1}^{T - 1} \mathbb{1}((x_j^{t+1} - x_i^t) \in \left[\phi_1, \phi_2\right])
\end{align}
where $\phi_1$ and $\phi_2$ are parameters of the heuristic.

Two issues may be spotted. First, this heuristic is not able to distinguish a
directed relationships if both neurons are activated in the same time step and
may wrongly characterize indirect relationships because of the time
resolution. Second, it may be necessary to take into account $s_{ji}$ to
estimate the quality of the pairwise coefficient $s_{i,j}$. Indeed, if both
coefficients are similar ($s_{ij} \sim s_{ji}$) the directivity can not be
inferred since the order of fluorescent rising is not well defined. On the
contrary, different coefficients $s_{ij}$ and $s_{ji}$ mean that an order of
fluorescent rising is well established and observable. There we choose to
compute a differential pairwise coefficient $z_{ij}$ defined by the following
equation
\begin{align}
z_{ij} = \sum_{t=1}^{T - 1}
    \mathbb{1}((x_j^{t+1} - x_i^t) \in \left[\phi_1, \phi_2 \right]) -
    \mathbb{1}((x_i^{t+1} - x_j^t) \in \left[\phi_1, \phi_2 \right]),
\end{align}
where $T$ is the number of time steps, $\phi_1$ and $\phi_2$ are parameters.

At the light of the poor results of this heuristic taken on its own
(on Table \ref{tab:directivity} in Section~\ref{sapp:results}) as \textit{directivity}, we use it to make
asymmetric the degree of association matrix using partial correlation
statistics. We expect that when the heuristic is highly confident (i.e.,
$z_{i,j} \gg 0 \gg z_{j,i}$) the degree of association   of edge $(i,j)$ is
increased while the degree of association corresponding to the edge $(j,i)$ is
reduced. Even if this adjustment is subtle, it might have a slight positive
effect on the ROC AUC.

@@

Theses filters lead, taken individually, to worse performances  than filters
$f_1$ and $f_2$ but averaging the results of all low-pass filters leads to better ROC AUC.
\begin{table}[ht]\label{tab:comparison}
%\caption{Evaluation of different methods on the four \textit{normal} datasets\label{tab:comparison}}
\centering
\small
\begin{tabular}{| l | c c c c | c c c c |}
\hline
& \multicolumn{4}{c|}{AUROC} & \multicolumn{4}{c|}{AUPRC} \\
\textit{Averaging} $\backslash$ \textit{normal-} & \textit{1} & \textit{2} & \textit{3} & \textit{4} & \textit{1} & \textit{2} & \textit{3} & \textit{4} \\
\hline
\hline
 with $f_1$, $f_2$ & 0.937 & 0.935 & 0.935 & 0.931 & 0.391 &  \textbf{0.390} &  0.385 & \textbf{0.375}  \\
 with $f_1$, $f_2$, $f_3$, $f_4$ & \textbf{0.938} & \textbf{0.936} & \textbf{0.936} & \textbf{0.932} & 0.391 & 0.389 & 0.385 & 0.374\\
\hline
\end{tabular}
\caption{Performance on \textit{normal-1,2,3,4} with partial correlation with different averaging approaches.}
\end{table}

\begin{table}[ht]\label{tab:comparison}
%\caption{Evaluation of different methods on the four \textit{normal} datasets\label{tab:comparison}}
\centering
\small
\begin{tabular}{| l | c c c c | c c c c |}
\hline
& \multicolumn{4}{c|}{AUROC} & \multicolumn{4}{c|}{AUPRC} \\
\textit{Full method} $\backslash$ \textit{normal-} & \textit{1} & \textit{2} & \textit{3} & \textit{4} & \textit{1} & \textit{2} & \textit{3} & \textit{4} \\
\hline
\hline
 without directivity & 0.943 & 0.942 & 0.942 & 0.939 & 0.403 & 0.404 & 0.398 & 0.388  \\
  with directivity  & \textbf{0.944} & \textbf{0.943} & 0.942 & \textbf{0.940} & \textbf{0.404} & \textbf{0.405} & \textbf{0.399} & \textbf{0.389}\\
\hline
\end{tabular}
\caption{Performance on \textit{normal-1,2,3,4} of ``Full Method'' with and without the directivity heuristic.}
\label{tab:directivity}
\end{table}

@@

We report here experiments on particular datasets provided by the organizers
which are similar to \textit{normal} datasets, i.e., made of 1000 neurons,
except for one parameter. We compute, as done in Section~\ref{sec:results},
both AUROC and AUPRC and results are reported in
Table~\ref{tab:results_appendix}. We will briefly recall what characterizes
each dataset:
\begin{description}
\item[lowcon]: Similar network but with a lower number of connections per neuron on average.
\item[highcon]: Similar network but with a higher number of connections per neuron on average.
\item[lowcc]: Similar network but with a lower clustering coefficient on average.
\item[highcc]: Similar network but with a higher clustering coefficient on average.
\item[normal-3-highrate]: Same topology as \textit{normal-3} but with a higher firing frequency, i.e., with highly active neurons.
\item[normal-4-lownoise]: Same topology as \textit{normal-4} but with a better signal to noise ratio.
\end{description}

>>>>

To filter out small variations in the signal obtained after applying the
function $g$ as well as to eliminate negative values, we use the following
hard-threshold filter
\begin{align}\label{eqn:hfilter}
h(x^{t}_i) &= x^{t}_i \mathbb{1}(x^{t}_i \geq \tau) \text{ with } \tau > 0,
\end{align}
yielding Figure \ref{fig:threshold_curve}.
As we can see, the processed signal only contains clean spikes.

@@

\section{Description  of the ``Full method''}
\label{app:optimized}

This section provides a detailed description of the method specifically tuned
for the Connectomics challenge. We will restrict our description to the
differences with respect to the simplified method presented in the main
paper. Most parameters were tuned so as to maximize AUROC on the
\textit{normal-1} dataset and our design choices were validated by monitoring
the AUROC obtained by the 145 entries we submitted during the
challenge. Although the tuned method performs better than the simplified one on
the challenge dataset and this tuning was probably decisive to get the best
performance, we believe that the tuned method clearly overfits the
simulator used to generate the challenge data and that the simplified method
should work equally well on new independent datasets. We nevertheless provide
the tuned method here for reference purpose. Our implementation of the tuned
method is available at \url{https://github.com/asutera/kaggle-connectomics}.

This appendix is structured as follows: Section~\ref{sapp:signal} describes the
differences in terms of signal processing. Section~\ref{sapp:averaging} then
provides a detailed presentation of the averaging
approach. Section~\ref{sapp:connectome} presents a heuristic to retrieve edge
directionality. Finally, Section~\ref{sapp:results} presents some experimental
results to validate the different steps of our proposal.

@@

In Section~\ref{sec:filter}, we introduce four filtering functions ($f$, $g$,
$h$, and $w$) that are composed in sequence (i.e., $w \circ h \circ g \circ
f$) to provide the signals from which to compute partial correlation
statistics.

Filtering is modified as follows in the tuned method:
\begin{itemize}
\item In addition to $f_1$ and $f_2$ (Equations \ref{eq:symetric-median} and
  \ref{eq:weighted-asymetric-median}), two alternative low-pass filters $f_3$
  and $f_4$ are considered:
\begin{align}
f_3(x^t_i) &= x^{t-1}_i + x^{t}_i + x^{t+1}_i + x^{t+2}_i, \label{eq:asymetric-median-forward} \\
f_4(x^t_i) &=  x_i^t + x^{t+1}_i  + x^{t+2}_i + x^{t+3}_i. \label{eq:asymetric-median}.
\end{align}
\item An additional filter $p$ is applied to smoothen differences in peak magnitudes
  that might remain after the application of the hard-threshold filter $h$:
\begin{align}
p(x^t_i) = (x_i^t)^c,
\end{align}
with $c=0.9$.
\item Filter $w$ is replaced by a more complex filter $w^*$ defined as:
\begin{align}
 w^*(x^{t}_i) &= {(x^{t}_i + 1 )^{\left (1 + \frac{1}{\sum_{j} x^{t}_j}\right )}}^{k(\sum_{j} x^{t}_j)}
\end{align}
where the function $k$ is a piecewise linear function optimized separately for
each filter $f_1$, $f_2$, $f_3$ and $f_4$ (see the implementation for full
details). Filter $w$ in the simplified method is a special case of $w^*$ with
$k(\sum_j x_j^t)=1$.
\end{itemize}
The pre-processed time-series are then obtained by the application of the
following function: $w^*\circ p \circ h \circ g \circ f_i$ (with $i=1$,2,3, or 4)

%% two composed filtering functions in
%% order (i) to remove noise induced by light scattering artifact, (ii) to take
%% into account the slow decay of fluorescence signal and (iii) to reduce the
%% importance of global high activity in the network. However, in order to win the
%% challenge, we have heavily optimized several different, but yet similar,
%% filtering composed function. On the one hand, each part of the composed
%% functions is optimized to maximize the ROC AUC (e.g., $w^*$). On the other
%% hand, we introduce new functions (e.g., $f_3$ and $f_4$) leading to alternative
%% composed functions. They might lead to worse ROC AUC when taken individually
%% but they improve the overall ROC AUC when taking as a part of the averaging
%% approach.

% Taken individually, those composed functions in
% conjunction with partial correlation might lead to worse AUROC, but improved
% whenever they are averaged.

%% Besides filters $f_1$ and $f_2$ respectively defined by equations \ref{eq:symetric-median} 
%% and \ref{eq:weighted-asymetric-median}, we also consider two
%% other low-pass filters
%% \begin{align}
%% f_3(x^t_i) &= x^{t-1}_i + x^{t}_i + x^{t+1}_i + x^{t+2}_i, \label{eq:asymetric-median-forward} \\
%% f_4(x^t_i) &=  x_i^t + x^{t+1}_i  + x^{t+2}_i + x^{t+3}_i. \label{eq:asymetric-median}
%% \end{align}

%% After the hard thresholded filter $h$, they were significant differences in
%% peak magnitudes. Since time series $x_i^t$  were mainly between 0 and 1,
%% we smoothened these differences in peak magnitudes through
%% \begin{align}
%% p(x^t_i) = (x_i^t)^c
%% \end{align}
%% yielding slightly better AUROC with $c=0.9 \in (0, 1)$.

%% After filter $p$, we applied a filter $w^*$ similar to $w$, whose goal is also
%% to magnify the importance of spikes that occur in case of low global activity.
%% Instead of equation \ref{eq:magnify-filter}, we had
%% \begin{align}
%%  w^*(x^{t}_i) &= {(x^{t}_i + 1 )^{\left (1 + \frac{1}{\sum_{j} x^{t}_j}\right )}}^{k(\sum_{j} x^{t}_j)}
%% \end{align}
%% where the function $k$ is a piecewise linear function optimized for each filter
%% $f_1$, $f_2$, $f_3$ and $f_4$. By comparison, the filter $w$ used the following
%% function $k(\sum_{j} x^{t}_j) = 1$. For more details on this aspect, we refer
%% the reader to the implementation \footnote{available on
%% \url{https://github.com/asutera/kaggle- connectomics}.}.

@@

As discussed in Section \ref{sec:inference}, the performance of the method (in
terms of AUROC) is sensitive to the value of the parameter $\tau$ of the
hard-threshold filter $h$ (see Equation \ref{eqn:hfilter}), and to the choice
of the low-pass filter (among $\{f_1, f_2, f_3, f_4\}$). Like in the simplified
method, we therefore averaged the partial correlation statistics obtained by
varying $\tau$ linearly from $0.100$ to $0.209$ (in step of $0.001)$ and by
varying the low-pass filter in $\{f_1, f_2, f_3, f_4\}$. Unlike in the
simplified method where each filter $f_1$ and $f_2$ were given equal weight in
the averaging, in the tuned method, we weighted each of the four filters
according to its individual performance on the \textit{normal-1}
dataset. \textcolor{red}{PG: je ne suis pas s\^ur de comprendre comment ces
  poids sont d\'etermin\'es}

%% Given the unequal AUROC performance of $f_1$, $f_2$, $f_3$ and $f_4$, the
%% contribution of each filter in the degree of association between two neurons
%% was chosen as to maximize the AUROC on \textit{normal-1} and is function of
%% their individual performance. By comparison in the simple method, each filter
%% $f_1$ and $f_2$ were given equal weight.

\subsection{Prediction of edge orientation}
\label{sapp:connectome}

Partial correlation statistics is a symmetric measure, while the connectome is
a directed graph. It could thus be beneficial to try to predict edge orientation. In
this section, we present a heuristic to retrieve directionality. %% Note however that
%% given the discussion at the end of Section~\ref{sec:results}, we cannot expect
%% a very large improvement in AUROC by predicting edge orientation.

This heuristic is based on the following observation. The rise of fluorescence
of a neuron indicates its activation. If another neuron is activated after a
slight delay, this could be a consequence of the activation of the first neuron
and therefore indicate a directed link in the connectome from the first to the
second neuron. Given this analysis, the idea of our heuristic is to count the
number of such events occurring between neurons $i$ and $j$ for consecutive
time steps. Mathematically, we compute for each pair $(i,j)$:
\begin{align}
s_{i,j} = \sum_{t=1}^{T - 1} \mathbb{1}((x_j^{t+1} - x_i^t) \in \left[\phi_1, \phi_2\right])
\end{align}
where $\phi_1$ and $\phi_2$ are parameters of the heuristic.

Two issues remain with this heuristic. First, it does not allow to detect a
directed relationship if both neurons are activated simultaneously at the same
time step, ie., when the delay between the two activations is below the time
resolution. Second, one needs to compare $s_{j,i}$ and $s_{i,j}$ to decide on
edge orientation. Indeed, edge orientation can only be predicted with some
confidence if either $s_{i,j}$ is greater than $s_{j,i}$ ($i\rightarrow j$) or
$s_{i,j}$ is lower than $s_{j,i}$ ($i\rightarrow j$). To address these issues,
we computed a differential pairwise coefficient $z_{ij}$ defined as follows:
\begin{align}
z_{ij} = \sum_{t=1}^{T - 1}
    \mathbb{1}((x_j^{t+1} - x_i^t) \in \left[\phi_1, \phi_2 \right]) -
    \mathbb{1}((x_i^{t+1} - x_j^t) \in \left[\phi_1, \phi_2 \right]),
\end{align}
where $T$ is the number of time steps, $\phi_1$ and $\phi_2$ are parameters.

\textcolor{red}{PG: je ne comprends pas ce qui est fait exactement. Mettre les
  maths? Valeurs de $\phi_1$ et $\phi_2$ ?} At the light of the poor results of this heuristic taken on its own
(on Table \ref{tab:directivity} in Section~\ref{sapp:results}) as
\textit{directivity}, we use it to make asymmetric the degree of association
matrix using partial correlation statistics. We expect that when the heuristic
is highly confident (i.e., $z_{i,j} \gg 0 \gg z_{j,i}$) the degree of
association of edge $(i,j)$ is increased while the degree of association
corresponding to the edge $(j,i)$ is reduced. Even if this adjustment is
subtle, it might have a slight positive effect on the ROC AUC.

@@

\paragraph{On the interest of low-pass filters $f_3$ and $f_4$.}

%%% SHOW DIFFERENTS COMPOSED FUNCTION

% Theses filters lead, taken individually, to worse performances  than filters
% $f_1$ and $f_2$ but averaging the results of all low-pass filters leads to better ROC AUC.

As reported in Table~\ref{tab:f3f4}, averaging over all low-pass filters leads
to better AUROC scores than averaging over only two low-pass filters, i.e., $f_1$ and
$f_2$. It however slightly deteriorates AUPRC.

\begin{table}[ht]
%\caption{Evaluation of different methods on the four \textit{normal} datasets\label{tab:comparison}}
\caption{Performance on \textit{normal-1,2,3,4} with partial correlation with different averaging approaches.}
\label{tab:f3f4}
\centering
\small
\begin{tabular}{| l | c c c c | c c c c |}
\hline
& \multicolumn{4}{c|}{AUROC} & \multicolumn{4}{c|}{AUPRC} \\
\textit{Averaging} $\backslash$ \textit{normal-} & \textit{1} & \textit{2} & \textit{3} & \textit{4} & \textit{1} & \textit{2} & \textit{3} & \textit{4} \\
\hline
\hline
 with $f_1$, $f_2$ & 0.937 & 0.935 & 0.935 & 0.931 & 0.391 &  \textbf{0.390} &  0.385 & \textbf{0.375}  \\
 with $f_1$, $f_2$, $f_3$, $f_4$ & \textbf{0.938} & \textbf{0.936} & \textbf{0.936} & \textbf{0.932} & 0.391 & 0.389 & 0.385 & 0.374\\
\hline
\end{tabular}
\end{table}

\paragraph{On the contribution of the directionality heuristic.}

Table~\ref{tab:dir} compares AUROC and AUPRC with and without prediction
of edge orientation. Both AUROC and AUPRC are (very slightly) improved by
directing the edges.

\begin{table}[ht]\label{tab:dir}
\caption{Performance on \textit{normal-1,2,3,4} of ``Full Method'' with and
  without prediction of edge orientation}
\label{tab:directivity}

%\caption{Evaluation of different methods on the four \textit{normal} datasets\label{tab:comparison}}
\centering
\small
\begin{tabular}{| l | c c c c | c c c c |}
\hline
& \multicolumn{4}{c|}{AUROC} & \multicolumn{4}{c|}{AUPRC} \\
\textit{Full method} $\backslash$ \textit{normal-} & \textit{1} & \textit{2} & \textit{3} & \textit{4} & \textit{1} & \textit{2} & \textit{3} & \textit{4} \\
\hline
\hline
 Undirected & 0.943 & 0.942 & 0.942 & 0.939 & 0.403 & 0.404 & 0.398 & 0.388  \\
  Directed & \textbf{0.944} & \textbf{0.943} & 0.942 & \textbf{0.940} & \textbf{0.404} & \textbf{0.405} & \textbf{0.399} & \textbf{0.389}\\
\hline
\end{tabular}
\end{table}

@@

We report in this apprendix the performance of the different methods compared
in the paper on 6 additional datasets provided by the challenge
organizers. These datasets, corresponding each to networks of 1000 neurons, are similar to
the \textit{normal} datasets except for one feature:
\begin{description}
\item[lowcon]: Similar network but with a lower number of connections per neuron on average.
\item[highcon]: Similar network but with a higher number of connections per neuron on average.
\item[lowcc]: Similar network but with a lower clustering coefficient on average.
\item[highcc]: Similar network but with a higher clustering coefficient on average.
\item[normal-3-highrate]: Same topology as \textit{normal-3} but with a higher firing frequency, i.e., with highly active neurons.
\item[normal-4-lownoise]: Same topology as \textit{normal-4} but with a better signal to noise ratio.
\end{description}

Results of several methods on these 6 datasets are provided in
Table~\ref{tab:results_appendix}. They confirm what we observed on the
\textit{normal} datasets. Average partial correlation and its tuned variant,
i.e.,``Full method'', clearly outperform other network inference methods on all
datasets. PC is close to GENIE3 and GTE but still slightly worse. GENIE3
performs better than GTE most of the times.
